MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-11 main/MSstatsPTM_Main.tex
Implementation (has not checked the figure caption yet)

The proposed methods are implemented in the open source R package $MSstatsPTM$. An overview of the steps of the package are illustrated in Figure \ref{fig:msstatsptm_design}. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. Conversion is done separately for the modified peptide and global profiling runs. To identify the local modification site we require the tool to identify where in the sequence the modification occurred. If the global profiling information is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. This amounts to testing the null hypothesis in~\eqref{conv_null_hyp}; which amounts to running the methods seen in $MSstats$ on modified peptide level data.

The next step is peptide/protein summarization. Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$. The package can optionally impute missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable. Finally, a separate model is fit to both the modified and global profiling runs. This model is automatically adjusted depending on the experimental design and acquisition method. If the base model is not applicable, a simplified model is fit. After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above.

We include visualizations for QC and model assessment. In QC plots the PTM and global protein are shown side by side, allowing for easier interpretation and comparison across data. In the model plots this is further separated into PTM, global protein, and adjusted PTM visualizations, allowing users to see the affect of the adjustment.

The package relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}. The statistical modeling relies on the functionality from the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}. The code is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. As expected when not using Empricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This was due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons. Additionally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and t-test

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
\caption{The workflow of $MSstatsPTM$ and how it fits into the experimental pipeline. The name of the R functions used for each step are highlighted in red. The package's workflow starts after modified and unmodified peptide quantification. First tool specific pre-processing is performed, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package. The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data. In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein. After both the summarization and group comparison steps, plots can be created to summarize the results.}
\label{fig:msstatsptm_design}
\end{figure}

>>>>

\todo{Could we try to use a consistent vocabulary for 'global protein' (we sometimes use unmodified peptides, unmodified features, global profiling etc.) Could you check the entire manuscript and see that we use the same terminology for that}

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. \todo{$MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as XXX and label-based data acquisitions such as XXX} $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.  \todo{Could you rephrase the next sentence to be more specific. If someone is a tool developer, what exactly should they do?} To identify the local modification site we require the tool to identify where in the sequence the modification occurred.  The package includes converters for seamless integration with multiple such tools. \todo{Could you clarify what the converters do, especially for PTM? Do we discard peptides with multiple modification sites?}. For experimental designs that include separate runs enriched in modified peptides and global profiling runs,  conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs. For experimental designs that include modified and unmodified peptides in a same run, \todo{what happens for the conversion?}

\todo{Could you rephrase the next sentence? What is summarized on the PTM side? The peptide sequence? Or any peptide sequence that contains the modification site?} The next step is peptide/protein summarization. Summarization is performed separately for the PTM and global protein in the same way as described above for $MSstats$. When summarizing the unmodified peptide features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable  results. Next, separate statistical models are fit to both the summaries of the modified features and the to summaries of the unmodified features. The models are automatically selected to reflect the experimental design and the data acquisition. If the base model is not applicable for a particular PTM or protein, e.g. due to missing data, a simplified model is fit.

After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above. Modification sites for which lacking the corresponding global profiling information cannot be adjusted for changes in protein abundance. In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide. In addition to the statistical functionalities, the implementation includes visualizations for quality control and assessment of the quality of model fit.

The implementation relies on functionalities from the R packages $MSstats$ \cite{Choi:2014} and $MSstatsTMT$ \cite{Huang:2020}, which in turn rely on the R packages $lme4$ \cite{Bates2015} and $lmerTest$ \cite{Kuznetsova2017}. $MSstatsPTM$ is available on Bioconductor, \url{http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}, and Github, \url{https://github.com/Vitek-Lab/MSstatsPTM}.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. As expected when not using Empiricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This was due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons. Additionally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and t-test

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
\caption{The $MSstatsPTM$ workflow. The names of the R functions used for each step are highlighted in red. The package's workflow starts after modified and unmodified peptide quantification. First tool specific pre-processing is performed, this includes modification site identification, general data cleaning, and formatting the data into the format needed for the package. The next step is feature level summarization, which summarizes features up to the modification level for the PTM data, and the protein level for the protein data. In the final step a model is fit to identify differential PTMs and unmodified proteins across conditions and the PTM model is adjusted for changes in the unmodified protein. After both the summarization and group comparison steps, plots can be created to summarize the results.}
\label{fig:msstatsptm_design}
\end{figure}
